---
title: "Reported Ethnicity and Predicted Ancestry"
author: "Simon Thompson"
output:
   html_document
---

```{r setup, include = FALSE}

rm(list = objects())
options(stringsAsFactors = FALSE,
    scipen = 200)
library(knitr)
library(wrangleR)
library(tidyverse)
library(DBI)
p <- getprofile(c("indx_con", "dams_con"))
indx_con <- dbConnect(RPostgres::Postgres(),
             dbname = "secondary_data_clean",
             host     = p$indx_con$host,
             port     = p$indx_con$port,
             user     = p$indx_con$user,
             password = p$indx_con$password)
dams_con <- dbConnect(RPostgres::Postgres(),
             dbname = "labkey",
             host     = p$dams_con$host,
             port     = p$dams_con$port,
             user     = p$dams_con$user,
             password = p$dams_con$password)

dt_ethnicity <- read.csv("dt_ethnicity.csv")

dams_d <- dbGetQuery(dams_con, "
select r.participant_identifiers_id as participant_id
    ,r.ethnicity_id
    ,ev.event_date::date as source_date
    ,'DAMS_CA' as source
from gelcancer.cancer_registration r
join gelcancer.event_detail ev 
    on r.event_details_id = ev.id
union
select r.participant_identifiers_id as participant_id
    ,r.ethnicity_id
    ,msg.metadata_date::date as source_date
    ,'DAMS_RD' as source
from rarediseases.rare_diseases_registration r
join rarediseases.rare_diseases_subject s 
    on r.subject_id = s.id 
join rarediseases.rare_diseases_message msg 
    on s.message_id = msg.id
;
") 

op_d <- dbGetQuery(indx_con, "
select participant_id
    ,ethnos as ethnicity_id
    ,apptdate::date as source_date
    ,'HES_OP' as source
from q4_19_nhsd.op
;") 

ae_d <- dbGetQuery(indx_con, "
select participant_id
    ,ethnos as ethnicity_id
    ,arrivaldate::date as source_date
    ,'HES_AE' as source
from q4_19_nhsd.ae
;")

apc_d <- dbGetQuery(indx_con, "
select participant_id
    ,ethnos as ethnicity_id
    ,admidate::date as source_date
    ,'HES_APC' as programme
from q4_19_nhsd.apc
where ethnos not in ('0', '1', '2', '3', '4', '5', '6', '7', '8', '9')
;")

d <- bind_rows(dams_d, op_d, ae_d, apc_d)
d <- merge(d, dt_ethnicity, by = "ethnicity_id", all.x = TRUE)

pa <- read.csv("/Users/simonthompson/Downloads/aggregate_gvcf_sample_stats.tsv",
    sep = "\t")
pa <- dropnrename(pa,
    c("participant_id",
    "pred_african_ancestries",
    "pred_south_asian_ancestries",
    "pred_east_asian_ancestries",
    "pred_european_ancestries",
    "pred_american_ancestries"),
    c("participant_id",
    "african",
    "south_asian",
    "east_asian",
    "european",
    "american"))
pa <- gather(pa, ancestry, prop, c("african", "south_asian", "east_asian", "european", "american"))

pretty_tab <- function(x) {
    as.data.frame(sort(table(x, useNA = "always"), decreasing = T))
}

pretty_prop_tab <- function(x) {
    as.data.frame(sort(prop.table(table(x, useNA = "always")), decreasing = T))
}

```

# Reported Ethnicity and Predicted Ancestry

Data used is reported ethnicity from DAMS and HES (OP, APC, AE).
Only those records using the new categories are included (70k records in APC are exclude).
The categories (and supercategories) used are:

```{r dt_ethnicity}

dt_ethnicity

```


## Overall counts of ethnicity categories

Below are the overall counts of different ethnicities and supercategories across all sources.

```{r overall_counts}
pretty_tab(d$ethnicity_label)
pretty_tab(d$ethnicity_supercat)
```

## Overall proportions of ethnicity categories

Below are the proportion of different ethnicities and supercategories across all sources.

```{r overall_prop}
pretty_prop_tab(d$ethnicity_label)
pretty_prop_tab(d$ethnicity_supercat)
```

## Number of different ethnicities per participant

The below plots show histograms of the number of different records per participant,
the number of different ethnicities pressent in those records (with and without the inclusion of Not Known/Not Stated).

```{r n_different_ethnicities}
n_eth <- d %>% 
    group_by(participant_id) %>%
    summarise(n_records = n(),
        count_exclude_not_known = n_distinct(ethnicity_id[!is.na(ethnicity_supercat)]),
        count_include_not_known = n_distinct(ethnicity_id)
    )

ggplot(n_eth, aes(x = n_records)) +
    geom_histogram(binwidth = 100) +
    labs(title = "Number of ethnicity records")
ggplot(n_eth, aes(x = count_include_not_known)) +
    geom_histogram(binwidth=1) +
    scale_x_continuous(breaks = c(0,2,4,8)) +
    labs(title = "Number of different ethnicities recorded per participant")
ggplot(n_eth, aes(x = count_exclude_not_known)) +
    geom_histogram(binwidth=1) +
    scale_x_continuous(breaks = c(0,2,4,8)) +
    labs(title = "Number of different ethnicities recorded per participant, excluding Not Known/Not Stated")

```

## Mode ethnicity per participant

Below are the counts of model ethnicity and supercategory across all participants.
Where there is a tie for a participant 'Unclear' is given.

```{r ethnicity_mode}

mode <- function(x) {
    t <- sort(table(x), decreasing = T)
    if (is.na(t[1])) {
        return(as.character(NA))
    } else if (length(t) == 1 | t[1] != t[2]) {
        return(names(t)[1])
    } else {
        return("Unclear")
    }
}

mode_eth <- d %>%
    group_by(participant_id) %>%
    summarise(mode = mode(ethnicity_label))

pretty_tab(mode_eth$mode)

```

```{r supercat_mode}

mode_supercat <- d %>%
    group_by(participant_id) %>%
    summarise(mode = mode(ethnicity_supercat))

pretty_tab(mode_supercat$mode)

```

## Distribution of predicted ancestry proportions

Below is a histogram of the calculated proportions of ancestry populations across all
participants. Only non-zero values have been plotted.

```{r predicted_ancestry_prop_dist}

ggplot(pa[pa$prop > 0,], aes(x = prop)) +
    geom_histogram(bins = 100) + 
    facet_grid(ancestry ~ .) + 
    labs(title = "Distribution of non-zero predicted ancestry proportions")

```
